---
title: "Analysis of ToothGrowth data"
author: "Premkumar Bhaskal"
date: "Sunday, February 22, 2015"
output: html_document
---

### Analysis of the ToothGrowth Data

### Summary
This Exercise is to perform the analysis of the ToothGrowth data in R.
Basically the data is about the tooth growth as a affect of 2 supplements VC and OJ, with different doses of 0.5, 1.0 and 2.0.
We first explore the data using the exploratory data analysis techniques.
Then we perform two group hypothesis testing to compare the mean of the distribution of the ToothGrowth by two different supplements.

```{r loading libraries}
# required libraries
  library(ggplot2)
  library(knitr)
```
```{r setoptions, echo = FALSE}
# setting global options.
opts_chunk$set(fig.width = 12)
```

### Exploring the data
```{r exploring data}
data(ToothGrowth)
head(ToothGrowth)

nrow(ToothGrowth)
summary(ToothGrowth)

unique(ToothGrowth$supp)

unique(ToothGrowth[,c("supp", "dose")])
```


### tooth growth vs supplement doses.
``` {r plotting the data}
## finding the average length of tooth growth grouped by supplements and doses.
toothGrowthBySuppDoes = aggregate(len ~ supp + dose, data = ToothGrowth, mean)

## plot of average growth Vs supplement_n_does
# g <- ggplot(data = toothGrowthBySuppDoes, aes(x = paste(supp,"-",dose), y = len))
g <- ggplot(data = toothGrowthBySuppDoes, aes(x = dose, y = len))
g <- g + geom_bar(stat = "identity")
g <- g + facet_grid(supp ~ .)
g <- g + labs(x = "dose used", y = "average tooth length")
g <- g + ggtitle("bar plot of average tooth growth vs dose grouped by different suppplements used")
print(g)
```


### two group testing - Hypothesis testing
We are doing a two group hypothesis testing assuming the NULL HYPOTHESIS that mean of the distribution of tooth growth is same 
on use of the two supplements (VC and OJ).


#### Z (97.5th quantile), in our case, n = 10, therefore (degrees of freedom)df = 9, so 
```{r z_quantile}
Z_1minusAlpha  = qt(0.975, df = 9)
Z_1minusAlpha
```

#### testing supplement dose of 0.5
```{r dose1 testing}
dose1Data <- ToothGrowth[ToothGrowth$dose == 0.5,]
t_test <- t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = dose1Data)
```
Here we see that the test statistic `r round(t_test$statistic,2)` is greater than the 97.5 quantile (Z(1-alpha)) of 2.26. so we reject the NULL hypothesis.

#### testing supplement dose of 1.0
```{r dose2 testing}
dose2Data <- ToothGrowth[ToothGrowth$dose == 1.0,]
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = dose2Data)
```
Again here we see since test_statistic `r round(t_test$statistic,2)` is greater than the Z_quantile (97.5%) of 2.26, so we reject the NULL hypothesis.

#### testing supplement dose of 2.0
````{r dose3 testing}
dose3Data <- ToothGrowth[ToothGrowth$dose == 2,]
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = dose1Data)
```

finally since the absolute test Statistic `r round(t_test$statistic,2)` is lesser than Z_1minusAlpha, we FAIL to reject the NULL hypothesis, so the 2 data groups have same mean.

### Conclusion
By hypothesis testing, we conclude that ToothGrowth distribution of supplements when taken in dose of 2.0, have a similar mean.
