---
title: "Analysis of effects of major weather events in the United States"
author: "Premkumar Bhaskal"
date: "Saturday, January 24, 2015"
output: html_document
---

###  Summary
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm 
database. This database tracks characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
Specifically this analysis is to find out two things.
 1.  The events which are most harmful with respect to population health.
 2.  The events which have greatest economic consequences.

###  Data Processing
```{r loading libraries}
# required libraries
  library(dplyr)
  library(ggplot2)
  library(knitr)
```


```{r setoptions, echo = FALSE}
# setting global options.
opts_chunk$set(fig.width = 12)
```

####  Checking out the data
Since the csv file is very big, reading the first few lines to check the format of the data present.
```{r verifying_data}
# checking the data before loading.
  con <- file("StormData.csv.bz2")
  fileData <- readLines(con, 10)
  close(con)
```
  
The file is plain csv, so reading it directly. (note that read.csv() reads compressed file directly without 
the  need to uncompress)
```{r loading_csv, cache = TRUE}
  stormData <- read.csv("StormData.csv.bz2")
```
  
Once the file is loaded, checking the size of the data. and checking the different types of events present.
Since total events are quite large, it is better to plot only top events.
```{r checking_out_data}
# checking out the data, dimensions etc.
  dim(stormData)
# analyzing types of events
  length(unique(stormData$EVTYPE))
```

### Cleaning weather events.
  First finding the unique activities and then sorting them , then writing to a file and then analyzing it.
  After analysing the file we understand that weather event column has lot of inconsistencies. Here we try
  to clean up some of these. Some cleaning is done by using below steps.  
  1. Converting to upper case.  
  2. Replacing synonyms with common terms.  
  3. Correcting spelling mistakes.  
```{r clean_weather_event}
  stormData$EVTYPE <- toupper(stormData$EVTYPE)
  
  stormData$EVTYPE <- stormData$EVTYPE[order(stormData$EVTYPE)]
  write.table(stormData$EVTYPE, "weather_events.txt")
  
  stormData$EVTYPE <- gsub("WINDS", "WIND", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*HEAVY SNOW.*", "HEAVY SNOW", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*BLIZZARD.*", "HEAVY SNOW", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*TSTM.*", "THUNDERSTORM", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*THUNDERSTORM.*", "THUNDERSTORM", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("COLD.*", "COLD", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*DRY MICROBURST.*", "DRY MICROBURST", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*FLASH.*FLOOD.*", "FLOOD", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*HAIL.*", "HAIL", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("PRECI.*TION", "RAIN", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("RAINFALL", "RAIN", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*HEAVY.*RAIN.*", "RAIN", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*WATERSPOUT.*", "WATERSPOUT", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*THU.*TORM.*", "THUNDERSTORM", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("SNOW.*", "SNOW", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("ICE.*", "ICE", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*HURRICANE.*", "HURRICANE", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*HIGH.*WIND.*", "HIGH WIND", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*LIG.*ING.*", "LIGHTNING", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*FREEZING.*", "FREEZING", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("FLOOD.*", "FLOOD", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*TORNADO.*", "TORNADO", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*TORNDAO.*", "TORNADO", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*TROPICAL STORM.*", "TROPICAL STORM", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*HEAVY SURF.*", "HEAVY SURF", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("EXTREME", "EXCESSIVE", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*EXCESSIVE.*HEAT.*", "EXCESSIVE HEAT", stormData$EVTYPE)
  stormData$EVTYPE <- gsub(".*EXCESSIVE.*WIND.*", "EXCESSIVE WIND", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("DRY.*", "DRY", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("WINTER.*", "WINTER", stormData$EVTYPE)
  stormData$EVTYPE <- gsub("RIP CURRENT.*", "RIP CURRENT", stormData$EVTYPE)

```

The events causing more fatalities and injuries are treated separtely.  
First finding the total fatalities caused by each of the different event and storing the top 10 events.
```{r fatalities_analysis}
# finding events causing fatalities.
  fatalitiesVsEvtype <- aggregate(FATALITIES ~ EVTYPE, data = stormData, sum)
  fatalitiesVsEvtype <- fatalitiesVsEvtype[order(fatalitiesVsEvtype$FATALITIES, decreasing = TRUE), ]
# top events causing major fatalities
  topEventForFatalities <- head(fatalitiesVsEvtype, 10)
```

Next finding the total injuries caused by each of the different event and storing the top 10 events.
```{r injuries_analysis}
# finding events causing injuries.
  injuriesVsEvtype <- aggregate(INJURIES ~ EVTYPE, data = stormData, sum)
  injuriesVsEvtype <- injuriesVsEvtype[order(injuriesVsEvtype$INJURIES, decreasing = TRUE),]
# top events causing major injuries.
  topEventForInjuries <- head(injuriesVsEvtype, 10)
```

Finding the events causing financial damages.  
The damages are separated as crop damages and property damages. The damages are provided in 2 columns.
The value part in one of the column like (CROPDMG and PROPDMG) and the exponent part in other column (CROPDMGEXP and PROPDMGEXP). The exponent part could either be a number or it can be a character like K, M, B, denoting kilo, Mega and Billion etc.  
To calculate, value should be multiplied with exponent (raised to 10).  

To accomplish this, the exponent column in the data is converted to upper case. Then each exponent character is mapped to corresponding numerical value. All the other characters (like ?,-,, etc) are mapped to 0 (so that 10^0 == 1). A column is added for each of the property and crop damage exponent in the existing data viz. propexp and cropexp. Then a new column 'totalDamage' is added using the below formula  
  "totalDamage = propdmg * 10^propexp + cropdmg * 10^cropexp"  

Finally the sum of totalDamage caused by each of the different event is found and the top 10 events are stored.
```{r damage_analysis}
# convert all to upper case
  stormData$PROPDMGEXP <- toupper(stormData$PROPDMGEXP)
  stormData$CROPDMGEXP <- toupper(stormData$CROPDMGEXP)
# convert non character exponent to 0
  stormData[!grepl("[K|M|H|B|0-9]", stormData$PROPDMGEXP), "PROPDMGEXP"] <- 0
  stormData[!grepl("[K|M|H|B|0-9]", stormData$CROPDMGEXP), "CROPDMGEXP"] <- 0
  
# map for character vs exponent value.
  char <- c(0,1,2,3,4,5,6,7,8,9,"H","K","M","B")  
  val <- c(0,1,2,3,4,5,6,7,8,9,2,3,6,9)
  expVsVal <- data.frame(char, val)
  
# merging with original data to add two new columns propexp and cropexp.
  colnames(expVsVal) <- c("char", "propexp")
  stormData <- merge(x = stormData, y = expVsVal, by.x = "PROPDMGEXP", by.y = "char")
  
  colnames(expVsVal) <- c("char", "cropexp")
  stormData <- merge(x = stormData, y = expVsVal, by.x = "CROPDMGEXP", by.y = "char")
  
# adding a new column totaldamage, which is sum of property damage and crop damage.
  stormData <- mutate(stormData, totalDamage  = PROPDMG * 10 ^ propexp + CROPDMG * 10 ^ cropexp)
  
# finding the sum of totalDamage for each of the different event.
  totalDamageVsEvtType <- aggregate(totalDamage ~ EVTYPE, data = stormData, sum)

# getting the events causing maximum damage
  totalDamageVsEvtType <- totalDamageVsEvtType[order(totalDamageVsEvtType$totalDamage, decreasing = TRUE), ]
  
# dividing by billion to fit the scale
  totalDamageVsEvtType$totalDamage <- totalDamageVsEvtType$totalDamage / 10^9

# top events causing financial damages
  topEventsForDamage <- head(totalDamageVsEvtType, 10)
  
```

###  Results
Below results depict the events reponsible for maximum fatalities, injuries and economical damage.
The table show the top 10 events, followed by a bar plot comparing the different events.
``` {r results}
  kable(topEventForFatalities, caption = "top events causing fatalities", 
        format = "markdown", row.names = FALSE)
  
  g <- ggplot(data = topEventForFatalities, aes(x = EVTYPE, y = FATALITIES))
  g <- g + geom_bar(stat = "identity")
  g <- g + labs(x = "event", y = "total fatalities")
  g <- g + ggtitle("top events causing fatalities")
  g <- g + theme(axis.text.x = element_text(angle = 90))
  print(g)
  
  kable(topEventForInjuries, caption = "top events causing injuries",
        format = "markdown", row.names = FALSE)
  
  g <- ggplot(data = topEventForInjuries, aes(x = EVTYPE, y = INJURIES))
  g <- g + geom_bar(stat = "identity")
  g <- g + labs(x = "event", y = "total injuries")
  g <- g + ggtitle("top events causing injuries")
  g <- g + theme(axis.text.x = element_text(angle = 90))
  print(g)
  
  kable(topEventsForDamage, caption = "top events causing damages",
        format = "markdown", row.names = FALSE)
  
  g <- ggplot(data = topEventsForDamage, aes(x = EVTYPE, y = totalDamage))
  g <- g + geom_bar(stat = "identity")
  g <- g + labs(x = "event", y = "total damages in Billion Dollars")
  g <- g + ggtitle("top events causing financial damages")
  g <- g + theme(axis.text.x = element_text(angle = 90))
  print(g)
```


